### 2022-06-15 paper.tex
implement Andre's suggestions

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is usually to assign cell types or states to cells. To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

How should one obtain the ensemble average (the curved line in Fig.\ \ref{fig:smoothres}B)? A simple approach to get a value for a specific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated. However, especially when only few cells offer coverage, these averages will be very noisy. Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood. The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; we got good results with 1000~bp and used this value for the examples presented here.

@@

In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each such tiles. Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.

Therefore, we propose the following approach (Fig. \ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size. Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{fig:smoothres}B. Then, calculate for each window the variance of these values over all cells. Select, say, the top 2\% windows with the highest variances and mark them as VMRs. Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR. Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.

@@

To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.

>>>>

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is usually to assign cell types or states to cells. To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on a large number of genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

How should one obtain the ensemble average (the curved line in Fig.\ \ref{fig:smoothres}B)? A simple approach to get a value for a specific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated. However, especially when only few cells offer coverage, these averages will be very noisy. Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood. The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; for the examples presented here we used 1000~bp.

@@

In the standard approach, one divides up (tiles) each chromosome into non-overlapping, equal-sized intervals, and quantitates the methylation of each tile. Such rigid placing of interval boundaries is unlikely to be optimal: for example, a VMR might be much smaller than a tile and its signal will hence be drowned out by the larger part of uninformative CpGs that are equal in all cells, when averaging over all the CpG sites in the tile.

Therefore, we propose the following approach (Fig. \ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size. Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{fig:smoothres}B. Then, calculate for each window the variance of these values over all cells. Select, say, the top 2\% windows with the highest variances and mark them as VMRs. Wherever thus marked windows overlap or are separated by only a small gap, merge them into one larger VMR. Then, calculate for each of these larger VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpG sites.

@@

To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression (Figure \ref{figure:correlation}C).
